---
title: "Covid19_Data"
author: "K. McQuinn"
date: "6/10/2022"
output: html_document
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE) 
```

Optional package install, these packages will not be automatically installed on your machine but will be used for this analysis.

``` {r install_packages, echo = TRUE, eval = FALSE}
install.packages("tidyverse")
install.packages("dplyr")
install.packages("plyr")
install.packages("lubridate")
install.packages("ggplot2")
install.packages("treemapify")
```

Reading in necessary libraries:

```{r library_packages, echo = TRUE, eval = TRUE}
library(tidyverse)
library(dplyr)
library(plyr)
library(lubridate)
library(ggplot2)
library(treemapify)
```

Reading in global COVID-19 data:

```{r GlobalData read_in, echo = TRUE}
#open data from JHU Github 

GlobalCases_URL <-"https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv"
GlobalDeaths_URL <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv"

GlobalCases_raw <- read.csv(GlobalCases_URL, na.strings = c(""," ","NA"))
GlobalDeaths_raw <- read.csv(GlobalDeaths_URL, na.strings = c(""," ","NA"))
```

Cleaning global data:

```{r Clean Global Data, echo = TRUE}
#pivot GlobalCases to give date just one column
GlobalCases <- GlobalCases_raw %>% pivot_longer (cols = -c('Province.State', 'Country.Region', Lat, Long), names_to = "Date", values_to = "Cases") %>% select (-c(Lat,Long))

#pivot GlobalDeaths to give date just one column
GlobalDeaths <- GlobalDeaths_raw %>% pivot_longer (cols = -c('Province.State', 'Country.Region', Lat, Long), names_to = "Date", values_to = "Deaths") %>% select (-c(Lat,Long))

#Joining Case data with global data and making new date column
Global <- GlobalCases %>%
  full_join(GlobalDeaths) %>%
  mutate (Date = mdy(substr(Date,2,9)))

#keep only columns that matter to my analysis
Global <- Global[ , c("Province.State", "Country.Region", "Date", "Cases", "Deaths")] %>%
  filter(Cases > 0)

#Rename columns for ease of use
names(Global)[1] <- "Province_State"
names(Global)[2] <- "Country_Region"

Global <- Global[ , c("Province_State", "Country_Region", "Date", "Cases", "Deaths")]
```

Adding population data to Global Data

```{r Read in Population Data, echo = TRUE}

#Reading in population data
UID_URL <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
UID <- read.csv(UID_URL, na.strings = c(""," ","NA"))

```

```{r Add Population Data to Global, echo = TRUE}

#joining population data to Global dataframe
Global <- Global %>% 
  left_join(UID, by = c("Province_State", "Country_Region")) %>%
  select (Province_State, Country_Region, Date, Cases, Deaths, Population)

#Adding combined_key to match US data
Global <- Global %>%
     unite("Combined_Key",
           c(Province_State, Country_Region),
           sep = ", ",
           na.rm = TRUE ,
           remove = FALSE)

#Remove datapoints that have NA as population size
Global <- Global %>% drop_na(Population)

#Reordering Global columns for organization's sake
Global <- Global[ , c("Date","Combined_Key", "Province_State", "Country_Region", "Population", "Cases", "Deaths")]

```

Next, I wanted to perform a similar analysis on the Global COVID19 Data:

```{r Add new columns Global cases and  deaths, echo = TRUE}

#Grouping data by state, combining all country data
GlobalByCountry <- Global %>%
  group_by(Country_Region, Date) %>%
  summarize(Cases = sum(Cases), Deaths = sum(Deaths), Population = sum(Population)) %>%
  mutate(DeathsPerMil = Deaths * 1000000/Population) %>%
  select(Country_Region, Date, Cases, Deaths, DeathsPerMil, Population) %>%
  ungroup()

#Grouping data by date
GlobalByDate <- Global %>%
  group_by (Date) %>%
  summarize (Cases = sum(Cases), Deaths = sum(Deaths), Population = sum(Population)) %>%
  mutate (DeathsPerMil = Deaths * 1000000/Population) %>%
  select(Date, Cases, Deaths, DeathsPerMil, Population) %>%
  ungroup()
```

Initial visualization of Global COVID cases vs Global COVID deaths:

``` {r Visualize GlobalByDate, echo = TRUE}

GlobalByDate %>%
  filter(Cases>0) %>%
  ggplot(aes(x=Date,y=Cases)) +
  geom_line(aes(color="Cases"))+
  geom_point(aes(color="Cases"))+
  geom_line(aes(y=Deaths, color="Deaths"))+
  geom_point(aes(y=Deaths, color="Deaths")) +
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x=element_text(angle=90)) +
  labs(title = "Global COVID19", y = NULL)

```

Visualize Cases vs Deaths by a chosen Country: 

``` {r Visualize by Country, echo = TRUE}

#Define whichever country we are interested in
Country <- "Peru"

GlobalByCountry %>%
  filter(Country_Region == Country) %>%
  filter(Cases>0) %>%
  ggplot(aes(x=Date,y=Cases)) +
  geom_line(aes(color="Cases"))+
  geom_point(aes(color="Cases"))+
  geom_line(aes(y=Deaths, color="Deaths"))+
  geom_point(aes(y=Deaths, color="Deaths")) +
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x=element_text(angle=90)) +
  labs(title = str_c("COVID19 in ", Country), y = NULL)

```

Adding new Columns: NewCases & NewDeaths to GlobalByCountry & GlobalByDate:

``` {r Add new columns Global new cases and new deaths, echo = TRUE}

#Adding column that shows new cases and new deaths/day
GlobalByCountry <- GlobalByCountry %>% 
  mutate(NewCases = Cases-lag(Cases),
         NewDeaths = Deaths-lag(Deaths))

#Adding column that shows new cases and new deaths/day
GlobalByDate <- GlobalByDate %>%
  mutate(NewCases = Cases-lag(Cases),
         NewDeaths = Deaths-lag(Deaths))

```

Visualize Global New Cases vs New Cases/Deaths: 

``` {r Visualize by Global New Cases/Deaths, echo = TRUE}

GlobalByDate %>%
  ggplot(aes(x=Date,y=NewCases)) +
  geom_line(aes(color="NewCases"))+
  geom_point(aes(color="NewCases"))+
  geom_line(aes(y=NewDeaths, color="NewDeaths"))+
  geom_point(aes(y=NewDeaths, color="NewDeaths")) +
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x=element_text(angle=90)) +
  labs(title = "New COVID19 Cases & Deaths, Globally", y = NULL)

```

Visualize New Cases vs New Deaths by Country: 

``` {r Visualize New by Country, echo = TRUE}
#Define whatever country we are interested in
Country <- "India"

GlobalByCountry %>%
  filter(Country_Region == Country) %>%
  filter(Cases>0) %>%
  ggplot(aes(x=Date,y=NewCases)) +
  geom_line(aes(color="NewCases"))+
  geom_point(aes(color="NewCases"))+
  geom_line(aes(y=NewDeaths, color="NewDeaths"))+
  geom_point(aes(y=NewDeaths, color="NewDeaths")) +
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x=element_text(angle=90)) +
  labs(title = str_c("New Cases & Deaths from COVID19 in ", Country), y = NULL)

```

Analyzing data to find which countries were the "worst" in case and death numbers

``` {r Summarize all countries, echo = TRUE}

#add columns that show Cases and Deaths per thousand people
GlobalByCountryTotals <- GlobalByCountry %>%
  group_by(Country_Region) %>%
  summarize(Deaths = max(Deaths), Cases = max(Cases), Population = max(Population), 
            CasesPerThou = 1000*Cases/Population,
            DeathsPerThou = 1000*Deaths/Population) %>%
  filter(Cases>0, Population>0)

#Show 5 countries with highest death rate
GlobalByCountryTotals %>%
  slice_max(DeathsPerThou, n=10)

#Show 5 countrie with lowest death rate
GlobalByCountryTotals %>%
  slice_min(DeathsPerThou, n=10)

```

Now I will create a linear model to predict Global Deaths 

``` {r Global Linear Model, echo = TRUE}

#Create model to relate CasesPerThou and DeathsPerThou
GlobalModel <- lm(DeathsPerThou ~ CasesPerThou, data = GlobalByCountryTotals)

#Add column to dataframe that has predicted death rates
GlblByCntryTotPred <- GlobalByCountryTotals %>% 
  mutate(PredDeaths = predict(GlobalModel))

#Plot actual death rates vs predicted death rates
GlblByCntryTotPred %>% ggplot()+
  geom_point(aes(x=CasesPerThou, y=DeathsPerThou), color = "blue") +
  geom_point(aes(x=CasesPerThou, y=PredDeaths), color = "red") +
  labs(title = "Predicted Deaths vs Actual Deaths, Globally", y = NULL)

```

After observing this chart and seeing that the death rate prediction doesn't seem to be an excellent model of the acutal death rate, I wanted to add GDP per capita data to see if any given country's wealth has an effect on its death rate:

``` {r Read in GDP per Capita Data, echo=TRUE}

#Reading in GDP per capita data
GDPPerCapita_URL <- "https://raw.githubusercontent.com/ktmcquinn/COVID19_Data/main/GDP%20per%20Capita%20by%20Country.csv"
GDPPerCapita <- read.csv(GDPPerCapita_URL)

#Renaming Country.Region
names(GDPPerCapita)[1] <- "Country_Region"

```

``` {r Add GDP per Capita Data to GlblByCntryTotPred, echo=TRUE}

#Join GDP Data with existing Global Data
GlobalPredGDP <- GlblByCntryTotPred %>% 
  left_join(GDPPerCapita, by = c("Country_Region")) %>%
  select (Country_Region, Deaths, Cases, Population, CasesPerThou, DeathsPerThou, GDP, PredDeaths)

```

This graph shows each country plotted based on their GDP Per Capita and their reported COVID Deaths:

``` {r Plotting GDP Capita Data , echo=TRUE}

GlobalPredGDP %>% ggplot()+
  geom_point(aes(x=GDP, y=DeathsPerThou), color = "darkgreen") +
  labs(title = "GDP Per Capita vs COVID Deaths, Globally")+
  scale_y_log10()+
  scale_x_log10()

```

The above chart is interesting because, as a group, the data seem to have a slight positive curve. I.e. The higher the relative GDP per capita, the higher the COVID death rate. This is not exactly what I expected. I am curious now about how population density affects the COVID death rate. 

I will now study the data in COVID deaths in the US:

```{r USData read_in, echo = TRUE}

#open US COVID-19 data from JHU Github 
USCases_URL <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv"
USDeaths_URL <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv"

USCases_raw <- read.csv(USCases_URL, na.strings = c(""," ","NA"))
USDeaths_raw <- read.csv(USDeaths_URL, na.strings = c(""," ","NA"))

```

Cleaning US data:

```{r Clean US Data, echo = TRUE}

#pivot USCases to give date just one column
USCases <- USCases_raw %>% 
  pivot_longer (cols = -c(UID:Combined_Key), 
                   names_to = "Date", 
                   values_to = "Cases") %>% 
     select (-c(UID:FIPS, Lat,Long_))

#pivot USDeaths to give date just one column
USDeaths <- USDeaths_raw %>% 
  pivot_longer (cols = -c(UID:Population), 
                   names_to = "Date", 
                   values_to = "Deaths") %>% 
     select (-c(UID:FIPS, Lat,Long_))

#Joining Case data with global data and making new date column
US <- USCases %>%
  full_join(USDeaths) %>%
  mutate (Date = mdy(substr(Date,2,9)))

#keep only columns that matter to my analysis
US <- US[ , c("Admin2", "Province_State", "Country_Region", "Combined_Key", "Date", "Population", "Cases", "Deaths")] %>%
  filter(Cases > 0)

#Renaming Admin2
names(US)[1] <- "County"

#Reordering US
US <- US[ , c("Date","Combined_Key", "County", "Province_State", "Country_Region", "Population", "Cases", "Deaths")]

```

Here I start analyzing US Data:

```{r Add USByState, echo = TRUE}

#Grouping data by state, combining all county data
USByState <- US %>%
  group_by(Province_State, Country_Region, Date) %>%
  summarize(Cases = sum(Cases), Deaths = sum(Deaths), Population = sum(Population)) %>%
  mutate(DeathsPerMil = Deaths * 1000000/Population) %>%
  select(Country_Region, Province_State, Date, Cases, Deaths, DeathsPerMil, Population) %>%
  ungroup()

#Grouping data by county, combining all date data
USByCounty <- US %>%
  group_by(Province_State, Country_Region, County, Combined_Key) %>%
  summarize(Cases = sum(Cases), Deaths = sum(Deaths), Population = sum(Population)) %>%
  mutate(DeathsPerMil = Deaths * 1000000/Population) %>%
  select(Country_Region, Province_State, County, Combined_Key, Cases, Deaths, DeathsPerMil, Population) %>%
  ungroup()


#Grouping data by date
USByDate <- USByState %>%
  group_by(Country_Region, Date) %>%
  summarize (Cases = sum(Cases), Deaths = sum(Deaths), Population = sum(Population)) %>%
  mutate (DeathsPerMil = Deaths * 1000000/Population) %>%
  select(Country_Region, Date, Cases, Deaths, DeathsPerMil, Population) %>%
  ungroup()

```

Initial visualization of US COVID cases vs US COVID deaths:

``` {r Visualize USByDate, echo = TRUE}

USByDate %>%
  filter(Cases>0) %>%
  ggplot(aes(x=Date,y=Cases)) +
  geom_line(aes(color="Cases"))+
  geom_point(aes(color="Cases"))+
  geom_line(aes(y=Deaths, color="Deaths"))+
  geom_point(aes(y=Deaths, color="Deaths")) +
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x=element_text(angle=90)) +
  labs(title = "COVID19 in US", y = NULL)

```

Visualizing COVID Cases and COVID deaths data over time for any chosen state:

``` {r visualizing cases vs deaths, echo = TRUE}
State <- "California"

USByState %>%
  filter(Province_State == State) %>%
  filter(Cases>0) %>%
  ggplot(aes(x=Date,y=Cases)) +
  geom_line(aes(color="Cases"))+
  geom_point(aes(color="Cases"))+
  geom_line(aes(y=Deaths, color="Deaths"))+
  geom_point(aes(y=Deaths, color="Deaths")) +
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x=element_text(angle=90)) +
  labs(title = str_c("COVID19 in ", State), y = NULL)
```

Analyzing data to find which states were the "worst" in case and death numbers

``` {r Summarize all states, echo = TRUE}

#add column for deaths and cases per thousand residents
USByStateTotals <- USByState %>%
  group_by(Province_State) %>%
  summarize(Deaths = max(Deaths), Cases = max(Cases), Population = max(Population), 
            CasesPerThou = 1000*Cases/Population,
            DeathsPerThou = 1000*Deaths/Population) %>%
  filter(Cases>0, Population>0)

#Show 5 states with highest death rate
USByStateTotals %>%
  slice_max(DeathsPerThou, n=5)

#Show 5 states with lowest death rate
USByStateTotals %>%
  slice_min(DeathsPerThou, n=5)

```

Visualize Death per Thousand for all states:

```{r DeathsPerThou in ALL states visual, echo = TRUE}

USByStateTotals %>%
  ggplot(aes(area = DeathsPerThou, fill = DeathsPerThou, label = paste(Province_State))) +
  geom_treemap() +
  geom_treemap_text(colour = "white",
                    place = "centre",
                    size = 15) +
  scale_fill_viridis_c()

```

Now we can dig deeper about which particular states we are curious. This next graph shows the data of COVID cases and COVID Deaths from two chosen states for comparison. I picked Hawaii and Mississippi because they are in the top 5 lowest death rate and top 5 lowest death rate, respectively.  

``` {r Visualize by State, echo = TRUE}
#Define whatever states we are interested in
State1 <- "Mississippi"
State2 <- "Hawaii"

#filter USByState data set to exclude 0 cases and include only the first specified state
State1_DF <- USByState %>%
  filter(Province_State == State1) %>%
  filter(Cases>0)
#filter USByState data set to exclude 0 cases and include only the second specified state
State2_DF <- USByState %>%
  filter(Province_State == State2) %>%
  filter(Cases>0)

ggplot() +
  geom_line(data = State1_DF, aes(x = Date, y = Deaths, color = State1), color = "darkgreen") +
  geom_point(data = State1_DF, aes(x = Date, y = Deaths, color = State1), color = "darkgreen") +
  geom_line(data = State2_DF, aes(x = Date, y = Deaths, color = State2), color = "purple") +
  geom_point(data = State2_DF, aes(x = Date, y = Deaths, color = State2), color = "purple") +
  scale_y_log10() +
  theme(legend.position = "bottom") +
  labs(title = str_c("COVID-19 Deaths in ", State1, " & ", State2))

```

Adding new Columns: NewCases & NewDeaths to USByState & USByDate:

``` {r Add new columns US new cases and new deaths, echo = TRUE}

#adding column to USByState and USByDate to include new case and new death data
USByState <- USByState %>% 
  mutate(NewCases = Cases-lag(Cases),
         NewDeaths = Deaths-lag(Deaths))
USByDate <- USByDate %>%
  mutate(NewCases = Cases-lag(Cases),
         NewDeaths = Deaths-lag(Deaths))

```

Visualize US New Cases vs New Cases/Deaths: 

``` {r Visualize by US New Cases/Deaths, echo = TRUE}

USByDate %>%
  ggplot(aes(x=Date,y=NewCases)) +
  geom_line(aes(color="NewCases"))+
  geom_point(aes(color="NewCases"))+
  geom_line(aes(y=NewDeaths, color="NewDeaths"))+
  geom_point(aes(y=NewDeaths, color="NewDeaths")) +
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x=element_text(angle=90)) +
  labs(title = str_c("COVID19 in US"), y = NULL)

```

Visualize New Cases vs New Deaths by California (or any other given state): 

``` {r Visualize New by State, echo = TRUE}

#Define whatever state we are interested in
State <- "California"

USByState %>%
  filter(Province_State == State) %>%
  filter(Cases>0) %>%
  ggplot(aes(x=Date,y=NewCases)) +
  geom_line(aes(color="NewCases"))+
  geom_point(aes(color="NewCases"))+
  geom_line(aes(y=NewDeaths, color="NewDeaths"))+
  geom_point(aes(y=NewDeaths, color="NewDeaths")) +
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x=element_text(angle=90)) +
  labs(title = str_c("New Cases & Deaths from COVID19 in ", State), y = NULL)

```


To further explore, I'm going to find population density by US county to try to make some more connections. 


Reading in County Population Density:

```{r Reading in county pop density data, echo = TRUE}

#reading in raw US county data
USCountyPopulationDensity_raw <- "https://raw.githubusercontent.com/ktmcquinn/COVID19_Data/main/USCounty_Population_Density.csv"
USCountyPopDen <- read.csv(USCountyPopulationDensity_raw)

```

Now, I will join the population density to the existing USByCounty data set:

```{r joining pop density data to USByCounty}

#joining population density data with US county data
USByCounty <- USByCounty %>% 
  left_join(USCountyPopDen, by = c("Country_Region", "Province_State", "County", "Combined_Key")) %>%
  select (Country_Region, Province_State, County, Combined_Key, Cases, Deaths, DeathsPerMil, Population, Population_Density)

```

I want to show population density vs deaths per million:

```{r plot deathspermil against popdensity}
USByCounty %>% ggplot()+
  geom_point(aes(x=Population_Density, y=DeathsPerMil), color = "blue", alpha = 0.4) +
  labs(title = "Population Density vs Deaths per Million, United States")+
  scale_y_log10()+
  scale_x_log10()
```

This is interesting again because it's not exactly what I expected! I am surprised to see that the death rate per million residents seem to skew slightly down. I assumed we'd see a higher rate of cases in more densely populated counties but that, generally, doesn't seem to be the case. If I were to push it further, I would probably go county by count and look into COVID-19 rules and policies to see if there was any connection. 